---
title: allan saint-maximin appreciation post
date: '2022-02-20'
tags: ["football"]
keywords: ["football", "r", "rstats", "newcastle united"]
description: he dribbles when he wants.
slug: saint-max-appreciation
images: 
  - "https://ewen.io/saint-max-appreciation/featured.jpeg"
draft: no
output: 
  hugodown::md_document:
    fig_asp: 1.1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE, message = FALSE, echo = FALSE, out.width = "640px"
  )

library(worldfootballR)
library(tidyverse)
library(forcats)
library(slider)
library(glue)
library(janitor)
library(gghighlight)
library(ggrepel)
library(pilot)
library(ggtext)

set_pilot_family("iA Writer Quattro S")
unhighlighted_col <- "#8d99ae"
highlight_col <- "#2b2d42"
asm_col <- "#ef233c"
today <- format(Sys.Date(), "%d %b %Y")

# get data
big5_possession <- fb_big5_advanced_season_stats(2022, "possession", "player")
big5_def <- fb_big5_advanced_season_stats(2022, "defense", "player")
nufc_matches <- get_team_match_results(
  c("https://fbref.com/en/squads/b2b47a98/2020-2021/Newcastle-United-Stats",
    "https://fbref.com/en/squads/b2b47a98/Newcastle-United-Stats"
    )
  )
```

For my sins, I am a Newcastle United fan. Watching them play has been hard graft for a decent while. Going deep in the UEFA Cup seems like a very quaint memory. There's reason for renewed sporting optimism, with a helping of geopolitical pessimism, now [the club is owned by Saudi Arabia](https://www.bbc.co.uk/sport/football/58826899) (for what it's worth, the previous owner Mike Ashley, has been [regularly accused of paying staff below minimum wage](https://www.theguardian.com/business/2020/jul/23/sports-direct-undercover-minimum-wage-mike-ashley), so the bar is unthinkably low). 

These sport~~wash~~ing revolutions typically involve waves of new players turning up to appease the new regime and expectations, inevitably taking places off of (mostly) technically limited members of the squad first. Longer term, this gentrification tends to be all-encompassing. I wanted to pay tribute to the player that has managed to light up countless diabolical performances across some forgettable seasons, while there's time left to enjoy more. 

First, a chart to convey the relative pain that comes with watching the current Newcastle team.

```{r nufc-xg, out.width="100%", fig.asp=0.9}
nufc_xg <- nufc_matches %>%
  clean_names() %>%
  dplyr::filter(
    comp == "Premier League", !is.na(x_g)
  ) %>%
  mutate(
    xg_rolling = slide_dbl(x_g, mean, .before = 6),
    xga_rolling = slide_dbl(x_ga, mean, .before = 6),
    xg_diff = xg_rolling - xga_rolling,
    gf = as.numeric(gf),
    ga = as.numeric(ga),
    g_rolling = slide_dbl(gf, mean, .before = 6),
    ga_rolling = slide_dbl(ga, mean, .before = 6),
    g_diff = g_rolling - ga_rolling,
    date = as.Date(date),
    gameweek = as.numeric(str_extract(round, "[0-9]+")),
    season = case_when(
      date <= as.Date("2021-05-23") ~ "2020-21",
      date >= as.Date("2021-08-15") ~ "2021-22"
    )
  )

manager_changes <- tribble(
  ~gameweek, ~season,
  9, "2021-22",
  11, "2021-22"
)

managers <- tribble(
  ~name, ~gameweek, ~season, ~angle, ~y,
  "Steve Bruce", 1, "2021-22", 0, 1.5,
  "Graeme Jones", 10, "2021-22", 90, 0.5,
  "Eddie Howe", 12, "2021-22", 0, 1.5
)

series_labels <- tibble(
  name = c("xG difference", "goal difference"),
  colour = c(asm_col, highlight_col),
  gameweek = c(12, 2),
  y = c(.25, 1.5),
  season = "2020-21"
)

ggplot(
  nufc_xg,
  aes(x = gameweek)
  ) +
  facet_wrap(~season, scales = "free_x") +
  # add the series
  geom_ribbon(
    aes(ymin = xg_diff, ymax = pmin(xg_diff, g_diff)),
    alpha = 0.3, fill = asm_col) +
  geom_ribbon(
    aes(ymin = g_diff, ymax = pmin(xg_diff, g_diff)),
    alpha = 0.3, fill = highlight_col) +
  geom_line(aes(y = xg_diff, colour = "xG difference")) +
  geom_line(aes(y = g_diff, colour = "goal difference")) +
  # add reference lines + text
  geom_hline(
    yintercept = 0, linetype = "dashed", colour = unhighlighted_col
    ) +
  geom_vline(
    aes(xintercept = gameweek), data = manager_changes,
    linetype = "dashed", colour = unhighlighted_col
    ) +
  geom_text(
    aes(label = name, angle = angle, y = y), data = managers,
    hjust = 0, family = "iA Writer Quattro S", fontface = "bold",
    size = 3
  ) +
  geom_text(
    aes(label = name, y = y), data = series_labels, hjust = 0,
    family = "iA Writer Quattro S", fontface = "bold",
    size = 3,
    colour = c(asm_col, highlight_col)
  ) +
  # theme stuff
  scale_color_manual(
    values = c(highlight_col, asm_col), name = NULL
    ) +
  labs(
    title = "This season, Newcastle have <span style = 'color:#ef233c;'>under-performed</span> relative to an already poor xG difference",
    subtitle = "6 game rolling goal and xG difference | English Premier League",
    x = "gameweek",
    y = "goal/xG difference",
    caption = glue(
      "Players with at least 900 minutes played.
      Source: FBref, {today}
      @ewen_"
    )
  ) +
  theme_pilot(grid = "h", legend_position = "none") +
  theme(
    plot.title = element_textbox_simple(
      padding = margin(0, 0, 5, 0), hjust = 0, lineheight = 1
    ),
    plot.title.position = "plot"
  )
```
Not only has the xG difference been mostly below zero for two consecutive seasons (i.e. no. of goals Newcastle should have scored, given the quality of chances created, minus the no. of goals their opposition should have scored), their actual goal difference has mostly lagged behind this as well. This is characteristic of a low-quality team.

So, why bother? Even in the darkest days of Sports Direct FC, there have been glimmers of joy to grab with both hands. Once upon a time, Alan Pardew might have treated us with a Hatem Ben Arfa cameo. Today, we are pacified by the one-man dribbling band that is Allan Saint-Maximin.

```{r dribble-carries, out.width="100%"}
pl_possession <- big5_possession %>%
  clean_names() %>%
  dplyr::filter(
    comp == "Premier League", mins_per_90 >= 10
  ) %>%
  mutate(
    succ_dribbles_90 = succ_dribbles / mins_per_90,
    att_dribbles_90 = att_dribbles / mins_per_90,
    prog_carries_90 = prog_carries / mins_per_90,
    is_amn = player == "Allan Saint-Maximin"
  )

ggplot(
  pl_possession,
  aes(x = succ_dribbles_90, y = prog_carries_90)
  ) +
  geom_point(
    shape = 21, alpha = .9, size = 3,
    aes(fill = factor(is_amn), colour = factor(is_amn))
  ) +
  gghighlight(
    succ_dribbles_90 > quantile(succ_dribbles_90, 0.95) |
      prog_carries_90 > quantile(prog_carries_90, 0.95),
    unhighlighted_params = list(
      colour = unhighlighted_col, fill = unhighlighted_col, alpha = 0.3
    )
  ) +
  geom_text_repel(
    aes(label = player), colour = highlight_col,
    data = filter(
      pl_possession,
      succ_dribbles_90 > quantile(succ_dribbles_90, 0.95) |
        prog_carries_90 > quantile(prog_carries_90, 0.95)
    ),
    size = 2.5, family = "iA Writer Quattro S"
  ) +
  scale_color_manual(values = c(highlight_col, asm_col), guide = "none") +
  scale_fill_manual(values = c(highlight_col, asm_col), guide = "none") +
  labs(
    title = "<span style = 'color:#ef233c;'>Saint-Maximin's</span> volume dribbling is in Adama Traoré territory",
    subtitle = "Successful dribbles and carries per 90 minutes | English Premier League 2021-22",
    x = "dribbles completed per 90",
    y = "progressive carries per 90",
    caption = glue(
      "Players with at least 900 minutes played.
      Source: FBref, {today}
      @ewen_"
      )
  ) +
  theme_pilot(subtitle_size = 11) +
  theme(
    plot.title = element_textbox_simple(
      size = 18, padding = margin(0, 0, 5, 0),
      hjust = 0
    ),
    plot.title.position = "plot"
  )

```

Given that Adama has now left the league, if you want to watch a jinking winger there is literally no competition. This is compounded by the fact that he plays for Newcastle, where the next most dribble-happy teammate is a way off. The "second guy" is further away than any other PL team.

```{r top-club-dribblers, out.width="100%"}
pl_possession %>%
  group_by(squad) %>%
  slice_max(att_dribbles_90, n = 2, with_ties = FALSE) %>%
  mutate(
    dribble_diff = att_dribbles_90 - lag(att_dribbles_90),
    top_dribbles_90 = lag(att_dribbles_90),
    top_player = lag(player)
    ) %>%
  dplyr::filter(!is.na(dribble_diff)) %>%
  select(
    squad, top_player, player, top_dribbles_90,
    att_dribbles_90, dribble_diff
    ) %>%
  ggplot(
    aes(
      y = reorder(squad, -dribble_diff)
    )
  ) +
  geom_linerange(
    aes(xmin = att_dribbles_90, xmax = top_dribbles_90),
    colour = highlight_col
    ) +
  geom_point(
    aes(x = att_dribbles_90),
    shape = 21, size = 4, colour = highlight_col, fill = "#ffffff"
    ) +
  geom_point(
    aes(x = top_dribbles_90),
    shape = 21, size = 4, colour = highlight_col, fill = "#ffffff"
    ) +
  geom_text(
    aes(label = player, x = att_dribbles_90),
    hjust = 1, nudge_x = -.2, size = 3,
    family = "iA Writer Quattro S", colour = unhighlighted_col
  ) +
  geom_text(
    aes(label = top_player, x = top_dribbles_90),
    hjust = 0, nudge_x = .2, size = 3,
    family = "iA Writer Quattro S", colour = unhighlighted_col
  ) +
  scale_x_continuous(limits = c(0, 10), expand = c(0, NA)) +
  labs(
    title = "Saint-Maximin's dribble attempts dwarf the club's #2 dribbler, Joe Willock",
    subtitle = glue(
      "Top two players per club for attempted dribbles per 90 minutes | English Premier League 2021-22"
      ),
    x = "dribbles attempted per 90",
    y = NULL,
    caption = glue(
      "Players with at least 900 minutes played.
      Source: FBref, {today}
      @ewen_"
    )
  ) +
  theme_pilot(grid = "v", subtitle_size = 11) +
  theme(
    plot.title = element_textbox_simple(
      size = 18, padding = margin(0, 0, 5, 0),
      hjust = 0
    ),
    plot.subtitle = element_textbox_simple(
      size = 11, padding = margin(0, 0, 5, 0)
    ),
    plot.title.position = "plot"
  )
  
```

How does he do it? Well, it turns out that Allan doesn't really do much of anything out of possession. There's only one fella that pressures the opposition less (guess who).

```{r pressures, out.width="100%"}
pl_def <- big5_def %>%
  clean_names() %>%
  dplyr::filter(
    comp == "Premier League", mins_per_90 >= 10,
    pos %in% c("FW", "FW,MF", "MF,FW", "MF"),
  ) %>%
  mutate(
    tackles_90 = tkl_tackles / mins_per_90,
    pressures_90 = press_pressures / mins_per_90,
    is_amn = player == "Allan Saint-Maximin"
  )

ggplot(
  pl_def, aes(x = tackles_90, y = pressures_90)
  ) +
  geom_point(
    shape = 21, alpha = .7, size = 3,
    aes(fill = factor(is_amn), colour = factor(is_amn))
  ) +
  gghighlight(
    squad == "Newcastle Utd",
    unhighlighted_params = list(
      colour = unhighlighted_col, fill = unhighlighted_col,
      alpha = 0.3
      )
    ) +
  geom_text_repel(
    aes(label = player),
    data = filter(
      pl_def,
      (
        tackles_90 < quantile(tackles_90, 0.01) |
          pressures_90 < quantile(pressures_90, 0.01)
      ) | squad == "Newcastle Utd"
    ),
    box.padding = 0.5, colour = highlight_col,
    family = "iA Writer Quattro S", size = 2.5
  ) +
  scale_color_manual(values = c(highlight_col, asm_col), guide = "none") +
  scale_fill_manual(values = c(highlight_col, asm_col), guide = "none") +
  labs(
    title = "Only Ronaldo pressures less than <span style = 'color:#ef233c;'>Saint-Maximin</span> in the Premier League",
    subtitle = glue(
      "Tackles and pressures applied per 90 minutes | English Premier League, 2021-22"
    ),
    x = "tackles per 90",
    y = "pressures per 90",
    caption = glue(
      "Midfielders/Forwards with at least 900 minutes played.
      Source: FBref, {today}
      @ewen_"
    )
  ) +
  theme_pilot(subtitle_size = 11) +
  theme(
    plot.title = element_textbox_simple(
      size = 18, padding = margin(0, 0, 5, 0),
      hjust = 0
    ),
    plot.title.position = "plot"
  )

```

tired: ASM is preserved for when Newcastle are in possession and need him to get the team up the field.

inspired: ASM doesn't need to defend because he is doing up defences like this. Never change.

```{r, eval=TRUE}
blogdown::shortcode("youtube", "dI6wSTgM9Zs")
```
